Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-12-02 paper/main.tex
update text with likelihood stuff

\subsubsection{Direct Position-Space Metric}
\label{subsec:direct-metric}
In the scenario where no cells divide during the observed interval, we can use our extraction
algorithm to obtain the position of the cells and then compare them directly to the cellular
skeletons obtained from the computational results.
Given $N_c$ cells with $N_v$ vertices in $d$ dimensions at $N_t$ time points, the total information
can be stored in an array $\nu_{i,j,k,l}$ of shape $(N_c,N_v,d,N_t)$.
Comparing the results of a simulation $\nu$ with experimental values $\tilde{\nu}$ is thus only a
matter of comparing two of these arrays via a suitable cost function~\cite{Wang2020}.
In our approach we utilize a \ac{ls} approach given by

\begin{equation}
    L_{\text{LS}}(\vec{x}, \vec{y})
        = \sum\limits_{i,j,k,l}|\nu_{i,j,k,l} - \tilde{\nu}_{i,j,k,l}|^2
\end{equation}

which minimizes the distance between the corresponding vertices at each time point.
One important detail is that the ordering of vertices must be identical in the numerical and
experimental values.

>>>>

\subsubsection{Direct Position-Space Metric}
\label{subsec:direct-metric}
In the scenario where no cells divide during the observed interval, we can use our extraction
algorithm to obtain the position of the cells and then compare them directly to the cellular
skeletons obtained from the computational results.
Given $N_c$ cells with $N_v$ vertices in $d$ dimensions at $N_t$ time points, the total information
can be stored in an array $\nu_{i,j,k,l}$ of shape $(N_c,N_v,d,N_t)$.
Comparing the results of a simulation $\nu$ with parameters $\theta$ with experimental values
$\tilde{\nu}$ is thus only a matter of comparing two of these arrays via a suitable cost
function~\cite{Wang2020}.
In our approach we utilize a \ac{ls} approach given by
\begin{equation}
    L_{\text{LS}}(\theta|\vec{x}, \vec{y})
        = \sum\limits_{ijkl}|\nu_{i,j,k,l} - \tilde{\nu}_{ijkl}|^2
\end{equation}
which minimizes the distance between the corresponding vertices at each time point.
One important detail is that the ordering of vertices must be identical in the numerical and
experimental values.

% This metric can be used to compare the model output to datapoints obtained from microscopic images.
A common technique to investigate identifiability questions and provide insights into confidence
levels of the obtained results is the
\textit{profile-likelihood}~\cite{Murphy2000,Kreutz2013,Tnsing2023}.
It is able to provide confidence intervals which are invariant under parameter
transformations~\cite{Meeker1995}.
It builds on the so-called \textit{log-likelihood} which propagates uncertainties $\sigma_{ijkl}$ in
the datapoints.
Under the assumption of a constant, known uncertainty, i.e. $\sigma_{ijkl}=\sigma$, we can rewrite
the log-likelihood in terms of $L_\text{LS}$ via
\begin{align}
    L(\theta) &= \sum\limits_{ijkl}\left(\left(\frac{\nu_{ijkl}-\tilde{\nu}_{ijkl}}{\sigma_{ijkl}}\right)^2
    + 2 \log(\sigma_{ijkl})
    +\text{const}\right)\\
    &= \frac{L_\text{LS}(\theta)}{\sigma^2} + \text{const}
\end{align}
where we absorbed the terms $\log(\sigma)$ in the sum into the constant term.
With this simplification, it is important to pick a value for $\sigma$ such that the real
uncertainty is overestimated.
Furthermore, the preceding section (\nameref{subsec:position-extraction-algorithm}) has shown that
the error distribution of the underlying data follows a normal distribution which justifies the core
assumption of this framework.
We conclude that minimizing, the least squares value $L_\text{LS}$ under constant $\sigma$ is
equivalent to minimizing the \textit{log-likelihood} $L$ which is again equivalent to optimizing the
likelihood that the chosen parameters represent our model.

The \textit{profile likelihood} $PL(\theta_i)$ scans a single parameter $\theta_i$ along its axis
while reoptimizing the remaining ones.
We can obtain confidence intervals via
\begin{equation}
    PL(\theta_i) - L(\theta) \leq \Delta_\alpha
\end{equation}
where the value $\Delta_\alpha$ is calculated from the $(1-\alpha)$ quantile of a $\chi^2$
distribution with one degree of freedom~\cite{Tnsing2023}.
